A thermodynamic view of dusty protoplanetary disks: axisymmetric stability

### 2016-10-19 linear.tex
re-written thermo interpret section
final changes before break

\section{Compressible streaming instability} 
In the special case of an unstratified disk (neglecting the vertical
component of the stellar gravity), we may also Fourier analyze in $z$
to obtain a local dispersion relation of the form  
$\sum_{j=1}^{5}c_j(k_x,k_z)\sigma^j = 0$, where $k_z$ is a real
vertical wavenumber. The coefficients $c_j$ can be read 
off Eq. \ref{streaming_dispersion} in  Appendix \ref{compressible_streaming}. 
There we also show that this dispersion relation reduces to that for
the streaming instability in the limit of incompressible gas and small
$\tstop$ \citep{youdin05a,jacquet11}.

We solve Eq. \ref{streaming_dispersion} for selected cases where the  
growth rates of the streaming instability have been 
calculated analytically from full two-fluid disk models, as well as
numerically from particle-gas simulations 
\citep[namely][]{youdin07,bai10b}. These authors use normalized
wavenumbers $K_x = \eta r k_x$ and similarly for $K_z$, where
\begin{align} 
  \eta \equiv -\frac{1}{2\rhog r\OmK^2}\frac{\p P}{\p r} = 
  \frac{1}{2\left(1-\tepsilon\right)}\frac{F_r}{r\OmK^2}, 
\end{align} 
measures the pressure offset of Keplerian rotation. We use
$\eta=0.05c_s/r\OmK$. They also parametrize the stopping time with
$\tau_\mathrm{s}=\tstop/(1-\tepsilon)$.

@@

\begin{deluxetable*}{llrrrrr}
  \tablecolumns{7}
  \tablecaption{Selected eigenfrequencies for the streaming
    instability. \label{si_compare}
  }
  \tablehead{
    \colhead{Mode} & 
    \colhead{$\tau_\mathrm{s}\OmK$} &
    \colhead{$\epsilon$} &
     \colhead{$K_x = K_z$} &
      \colhead{$\sigma/\OmK$ (two-fluid)} &
    \colhead{$\sigma/\OmK$ (one-fluid, Eq. \ref{streaming_dispersion})}   &
    \colhead{$\mathcal{W}$ (arbitrary units) } 
  }
\startdata
 linA, \cite{youdin07} &  $0.1$       & 3.0 & 30    & $-0.3480 +
 0.4190\ii$ & $0.3647 + 0.4251\ii$ & $0.026$\\

linB, \cite{youdin07} & $0.1$        &  0.2 & 6 & $0.4999 +
0.0154\ii$&   $-0.4999 + 0.0054\ii$  & $1.74$ \\

>>>>

\section{Compressible streaming instability}\label{si}
In the special case of an unstratified disk (neglecting the vertical
component of the stellar gravity), we may also Fourier analyze in $z$
to obtain algebraic dispersion relation of the form  
$\sum_{j=0}^{5}c_j(k_x,k_z)\sigma^j = 0$, where $k_z$ is a real
vertical wavenumber. The coefficients $c_j$ can be read 
off Eq. \ref{streaming_dispersion} in  Appendix \ref{compressible_streaming}. 
There we also show that this dispersion relation reduces to that for
the streaming instability in the limit of incompressible gas and small
$\tstop$ \citep{youdin05a,jacquet11}.

We solve Eq. \ref{streaming_dispersion} for selected cases where the  
growth rates of the streaming instability have been 
calculated analytically from full two-fluid disk models and measured 
from direct numerical particle-gas simulations 
\citep[namely][]{youdin07b,bai10b}. These authors use normalized
wavenumbers $K_x = \eta r k_x$ and similarly for $K_z$, where
\begin{align} 
  \eta \equiv -\frac{1}{2\rhog r\OmK^2}\frac{\p P}{\p r} = 
  \frac{1}{2\left(1-\tepsilon\right)}\frac{F_r}{r\OmK^2}, 
\end{align} 
measures the pressure offset of Keplerian rotation. We use
$\eta=0.05c_s/r\OmK$. They also define the stopping time as 
$\tau_\mathrm{s}=\tstop/(1-\tepsilon)$.

@@

Notice also there is better agreement between eigenfrequencies for
smaller $\mathcal{W}$, i.e. smaller phase pressure-density phase
lags. While this is generally achieved with decreasing $\tstop$, other
problem parameters may affect the magnitude of this phase lag.
For example, `linA' and `linB' have the same $\tstop$ but linA shows
better agreement with the one-fluid growth rate. This suggest that how
well dust and gas are effectively coupled is problem-dependent.

%"generalized stokes number" would include other problem parameters,
%e.g. wavenumber

\begin{deluxetable*}{llrrrrr}
  \tablecolumns{7}
  \tablecaption{Selected eigenfrequencies for the streaming
    instability. \label{si_compare}
  }
  \tablehead{
    \colhead{Mode} & 
    \colhead{$\tau_\mathrm{s}\OmK$} &
    \colhead{$\epsilon$} &
     \colhead{$K_x = K_z$} &
      \colhead{$\sigma/\OmK$ (two-fluid)} &
    \colhead{$\sigma/\OmK$ (one-fluid, Eq. \ref{streaming_dispersion})}   &
    \colhead{$\mathcal{W}$ (arbitrary units) } 
  }
\startdata
 linA, \cite{youdin07b} &  $0.1$       & 3.0 & 30    & $-0.3480 +
 0.4190\ii$ & $0.3647 + 0.4251\ii$ & $0.026$\\

linB, \cite{youdin07b} & $0.1$        &  0.2 & 6 & $0.4999 +
0.0154\ii$&   $-0.4999 + 0.0054\ii$  & $1.74$ \\

### 2016-10-19 model.tex
final changes before break

Now, quite generally, Eq. \ref{lin_mass_full}---\ref{lin_energy_full} 
imply the integrand of the numerator in Eq. \ref{thermal_instability} 
is 
\begin{align} 
  \imag \left(\delta \mathcal{C}
  \nabla\cdot\dd \bm{v}^*\right) = 
  -|\sigma|^2\imag \left(\Delta P 
  \frac{\Delta\rho^*}{\rho}\right). \label{pdv}
\end{align}
%This states that `PdV' work (the right-hand-side) is 
%provided by dust-drag friction, and can be seen explicitly as 
%follows. 
Then combining Eq. \ref{work_real}, \ref{pdv} and
\ref{thermal_instability} 
gives  
\begin{align}
s = \frac{1}{\mathcal{I}^2}\mathcal{W}, \label{growth_work}
\end{align}
where we have set $\sigma= - \omega$ in 
Eq. \ref{pdv}. Eq. \ref{growth_work} shows that if the average work
done is positive during a periodic oscillation, $\mathcal{W}>0$, then
then its amplitude would actually grow ($s>0$).

To see when/where positive work is done, 
consider a point in the volume. We set $\Delta\rho = 1$ here 
and take $\omega>0$ without loss of generality. (Thus the phase of
perturbations increase in time.) This point contributes to positive
work done if $-\omega \imag\left(\Delta
P\Delta\rho^*\right)>0$, which requires $\imag\left(\Delta
P\right)<0$. This corresponds to pressure lagging behind density since
$\imag\left(\Delta\rho\right)=0$.

\subsubsection{Physical picture of dust-drag instabilities}

@@

Fig. \ref{pdv_cartoon} gives a graphical demonstration that 
if gas pressure lags behind total density, then positive work done 
($\mathcal{W}>0$) because it leads to a clockwise path in the
`P-V' plane. Consider a fluid element at maximum density during an
oscillation cycle (point $A$), so that $D\rho/ D t = 0$. If dust-gas
drag provides an effective heating to the fluid element at $A$
(i.e. gas influx)  then it will experience increasing pressure, $DP/D
t>0$. It attains pressure maximum (point $B$) \emph{after} reaching
density maximum.    
From $A$ to $B$, pressure forces increase while the parcel is already
expanding. This corresponds to positive work done.

>>>>

Now, quite generally, Eq. \ref{lin_mass_full}---\ref{lin_energy_full} 
imply the integrand of the numerator in Eq. \ref{thermal_instability} 
is 
\begin{align} 
  \imag \left(\delta \mathcal{C}
  \nabla\cdot\dd \bm{v}^*\right) = 
  -|\sigma|^2\imag \left(\Delta P 
  \frac{\Delta\rho^*}{\rho}\right). \label{pdv}
\end{align}
%This states that `PdV' work (the right-hand-side) is 
%provided by dust-drag friction, and can be seen explicitly as 
%follows. 
Then combining Eq. \ref{work_real}, \ref{pdv} and
\ref{thermal_instability} 
gives  
\begin{align}
s = \frac{\mathcal{W}}{\mathcal{I}^2}, \label{growth_work}
\end{align}
where we have set $\sigma= - \omega$ in 
Eq. \ref{pdv}. Eq. \ref{growth_work} shows that if the average work
done is positive during a periodic oscillation, $\mathcal{W}>0$, then
then its amplitude would actually grow ($s>0$).

Under what condition is positive work is done? 
Consider a point in the volume where $\Delta\rho = 1$  
and take $\omega>0$ without loss of generality. (Thus the phase of
perturbations increase in time.) This point contributes to positive 
work done if $-\omega \imag\left(\Delta 
P\Delta\rho^*\right)>0$, which requires $\imag\left(\Delta
P\right)<0$. This corresponds to pressure lagging behind density since
$\imag\left(\Delta\rho\right)=0$.

\subsubsection{Physical picture of dust-gas drag instabilities}

@@

Fig. \ref{pdv_cartoon} gives a graphical demonstration that 
if gas pressure lags behind density, then positive work done 
($\mathcal{W}>0$) because it leads to a clockwise path in the
`P-V' plane. From $A$ to $B$, a fluid parcel is expanding to return to 
equilbrium, but pressure (i.e. restoring) forces is still
increasing. This over-compensation corresponds to positive work done. 
Similarly, from $C$ to $D$ the parcel is contracting but pressure 
(restoring) forces are still dropping, which allows the 
parcel's contraction to over-shoot its equilibrium density. The
overall positive work done leads to growth in the oscillation amplitude.

%Consider a fluid element at maximum density during an
%oscillation cycle (point $A$), so that $D\rho/ D t = 0$. If dust-gas
%drag provides an effective heating to the fluid element at $A$
%(i.e. gas influx)  then it will experience increasing pressure, $DP/D
%t>0$. It attains pressure maximum (point $B$) \emph{after} reaching
%density maximum.

When applied to dusty gas, the work done is attributed to finite 
dust-gas drag, which allows relative drift between gas and dust. 
A pressure-density lag of $A$ to $B$ is achieved if, 
after a region reaches maximum total density, its 
gas component is still increasing. This in turn implies a sufficiently
large particle flux \emph{out} of that region.
